### 2015-10-05 Manuscript/SW_early_life_telomeres.Rmd
HLD edits. Abstract and citation style changed to Evolution

---
output:
  word_document:
    reference_docx: myref.docx
bibliography: MyBib.bib
csl: JAE.csl

@@

a. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
b. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
c. Behavioural Ecology and Self-organization Group, Centre for Ecological and Evolutionary Studies, University of Groningen, Groningen, The Netherlands
d. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
e. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles

@@

###Abstract
1. Favourable conditions early in life can confer delayed fitness advantages. However, the mechanistic basis of these 'silver spoon effects' remains poorly understood.  Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early-life telomere shortening is correlated with reduced later-life survival. As telomeres are able to measure the impact of early life costs, they may provide a link between early-life environmental variation and later-life fitness.

2. The prevalence and strength of silver spoon effects has consequences for individual fitness and population growth rates. Understanding the physiological and ecological links between early life experiences and later life survival is therefore fundamental to our understanding of life-history evolution and population ecology.

3. We studied how telomere length and shortening link spatiotemporal variation in early-life conditions to survival in the Seychelles warbler (*Acrocephalus sechellensis*). The Seychelles warbler system is excellently suited to addressing this question, due to the availability of long-term longitudinal survival and senescence data unconfounded by dispersal. We combined this with cross-sectional and longitudinal telomere measurements taken from over 300 birds from multiple cohorts, and analysed how variation in the early life environment within and across cohorts affects telomere dynamics and survival.

4. We show that early-life telomere length and shortening rates vary markedly across cohorts, and that this variation is related to temporal fluctuations in food availability. Individuals born in breeding seasons with high food availability have reduced telomere shortening compared to individuals born in years with low food availability. Within cohorts, we found that telomere length in nestlings was negatively related to tarsus length (which likely represents nestling growth), and positively related to the number of helpers present in the natal territory. However, neither tarsus length nor the number of helpers was related to telomere length in fledged birds. Finally, we found that longer telomeres and reduced telomere shortening rates in early life were associated with increased survival later in life.

5. Our results show that telomeres can reveal complex hidden costs at the individual and cohort levels, and suggest that they may provide a link between early-life conditions and late-life survival in wild populations.

**Keywords:** Ageing; Life-history; Seychelles warbler; Telomeres

@@

###Introduction
Being subject to favourable environmental conditions during development and growth can confer fitness advantages later in life (so called 'silver spoon effects') [@Grafen1988; @Monaghan2008]. There is now evidence that adult fitness can be affected by a range of early-life experiences, including variation in habitat quality and resource availability [@Madsen2000; @VandePol2006; @Hayward2013], population density [@Nussey2007;@Douhard2013] and natural or anthropogenic environmental disturbance [@Reid2003; @Cartwright2014]. Understanding the causes and consequences of these early-life experiences is key to understanding many ecological and evolutionary  processes, including  patterns of natural and sexual selection,  population growth rates and even local extinction [@Coulson2001; @Roach2014]. Understanding silver spoon effects is therefore of central interest to ecologists, evolutionary biologists and conservationists.

While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness. Because the phenotypic consequences of an adverse environment can depend on an individual's initial  condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006], the later-life consequences of a good or poor start are expected to vary among individuals, populations and species. Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships between the early-life conditions that are measured and adult phenotypes. Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009]. In order to better understand the later-life consequences of early-life experiences, we therefore need to understand, or at least be able to measure, how the environment differentially affects individuals within a population.

A key process that has been associated with adult survival is the rate at which telomere repeats are lost. Telomeres are protective caps on the ends of chromosomes which shorten with age [@Monaghan2006], and in response to oxidative stress levels which can be elevated  due to environmental factors [@VonZglinicki2002]. When telomeres reach a critical shortness cells senescence  [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003]. These links between telomere length and senescence have inspired a great deal of recent research into telomere ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014]. While there is little direct evidence the relationship between telomere dynamics and survival is causal [@Barrett2011; @Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, retaining signatures of ecological stress that are otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].

There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life [e.g. @Frenck1998; @Haussmann2003], and that the extent of this telomere shortening is influenced by the conditions experienced early in life [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015]. Importantly, early-life telomere dynamics are related to both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and to other parameters such as cognition [@Nettle2015a]. However, the link between early-life conditions and late-life survival is limited to a few species, and thus poorly understood. Moreover, how early life telomere dynamics vary over spatial and temporal scales is not known.

The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent model system for studying senescence in the wild [reviewed in @Hammers2015]. Due to the isolated nature of the study population and intensive field monitoring, we have excellent survival data and sampling material spanning many years (see methods, below). Ecological conditions an population density on Cousin are variable over space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a]. Seychelles warblers remain on their natal territories for at least six months, and variation in natal territory quality has been linked to changes in levels of oxidative stress [@VandeCrommenacker2011a]. However, neither natal territory quality nor natal local density appear to be linked to early-life or adult survival [@Brouwer2006; @Hammers2013]. Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordinates) in the natal territory has been associated with increased survival later in life [@Brouwer2012]. Lastly, we have an established protocol for assessing absolute telomere length in this species [@Barrett2012], and telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost [@Barrett2013a]. Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced early in life, and to assess the later life consequences of early-life conditions.

@@

Telomere lengths were measured from a total of `r nrow(FlSA)+nrow(chicks)+nrow(Loss)` samples from `r nrow(FlSA)+nrow(chicks)` birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`. Of these birds, `r nrow(chicks)` were nestlings less than one month old (chicks), while the remaining `r nrow(FlSA)` were fledglings and subadults aged between two and twelve months (hereafter referred to as fledglings). For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with samples from individuals caught both in their first year and then again within two years.  We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay  [full details in @Barrett2012].

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. Our sampling regime covers `r max(juv$LayYear)-min(juv$LayYear)` years, and temporal variation in rainfall and food availability on Cousin over this period is expected to result in variation  in a range of social and environmental variables [@Brouwer2012]. This makes it difficult to disentangle how spatial and temporal processes differentially affect early-life telomere dynamics. To overcome this, we analysed how variation in early life conditions was related to variation in early-life telomere length separately among and within breeding seasons.

First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated average telomere length among all birds born in a given breeding season (i.e. a cohort), excluding cohorts where we had samples from $\leq$ 5 birds. We tested for a difference in telomere length among cohorts using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among cohorts is explained by temporal variation in insect abundance.

@@

For individuals with longitudinal data we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events. To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013]. Using this dataset we repeated the above analyses of among and within cohort telomere dynamics, replacing telomere length with telomere loss as the response variable.

Finally, we analysed the effects of telomere dynamics on survival, again by looking at variation both within and among cohorts. First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean telomere length and rate of loss of each cohort with the mean lifespan of all individuals in that cohort, using linear regression. For this analysis we excluded any recent breeding seasons where >10% of individuals in a cohort were still alive. We then used parametric survival analysis to test whether individual-level telomere length and rate of loss were related to to survival. Again, we mean-centred variables within cohorts to eliminate between-season temporal variation in telomeres and survival. The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values. As explanatory variables of survival we included early-life telomere length/loss (we ran a separate analysis for each, using the cross-sectional and longitudinal data, respectively), plus the factors identified as being related to telomere length in the previous analysis. With this approach we aimed to assess the effect of early-life telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere dynamics.

@@

#####Temporal variation in early life telomere dynamics
Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), and this variation in average telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig. 2A). Running the regression of cohort-level telomere length and food availability separately for each age class revealed a positive, significant relationship in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,5])`), but no relationship in nestlings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,5])`).

@@

#####Spatial variation in early life telomere dynamics
The top model explaining within-season variation in early-life telomere length contained age class, tarsus length and the number of helpers present in the natal territory, as well as interactions between age class and both tarsus length and the number of helpers (Table S1). This model was much better supported than the null model ($\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=juvall))-AICc(lm(cenTL~Fledged*(cenTarsus+cenHelper),data=juvall)))`). Both tarsus length and the number of helpers had high relative importance in the top model set (Fig. 3A), and the model containing these variables explained a reasonable amount of variation in telomere length (R^2^ = `r round2(summary(lm(cenTL~Ageclass*(cenTarsus+cenHelper),data=juvall))$r.squared)`). Telomere length was positively related to the number of helpers in nestlings, but not in fledglings (Fig. 3B). Tarsus length was negatively related to telomere length in nestlings, but not in fledglings (Fig. 3C).

For the subset of individuals with longitudinal data, we tested whether the factors previously identified as being related to early-life telomere length (i.e. tarsus length and number of helpers) were also associated with differences in rates of early life telomere loss. Due to a limited sample size, we considered all age classes together for this analysis. We found that none of these variables were associated with differences in telomere loss (all *P* > 0.05).

#####Early life telomere dynamics and survival
At the population-level, there was a positive but non-significant relationship between telomere length and lifespan (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig. 4A). Testing this relationship separately for each age class revealed that there was no age-specific relationship between cohort-level telomere length and lifespan (both *P* > 0.2). There was a negative relationship between population-level telomere loss and lifespan, although this was marginally non-significant (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig. 4B).

For the individual-based survival analysis, a log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution (Weibull) = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`). A survival model including all first year birds showed that early-life telomere length had a positive effect on survival, but this relationship was marginally non-significant (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`). Running this analysis among age classes revealed differential survival effects. Nestling telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; Fig. 5A), but telomere length at the fledgling stage was significantly related to survival (estimate = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`). This effect was positive, with longer telomeres in early life associated with increased survival later in life (Fig. 6B). Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3], lessthan = F)`).

@@

Our finding that food availability is related to early-life telomere length and rate of shortening adds to a growing body of evidence from humans and wild animals showing that the natal environment can have pronounced impacts on early-life telomere dynamics [reviewed in @Price2013; @Monaghan2014].  However, very few studies have shown that temporal variation in environmental conditions affects telomere dynamics in natural populations, and the studies that have done so are limited to just two seasons [@Mizutani2013; @Watson2015]. The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been possible in other studies, to show that temporal variation in environmental conditions does indeed affect telomere dynamics, and that this effect can be seen at the population level. Interestingly, in the Seychelles warbler annual variation in food availability is not directly linked to survival [@Brouwer2006; @Hammers2013]. It has been suggested that telomeres act as biomarkers of cost in natural populations, as they are able to capture individual and cohort-level variability in the costs experienced as a result of environmental pressures [@Asghar2015]. Our results confirm that telomeres can detect hidden costs in natural populations that are not detectable using life-history and environmental data alone

>>>>

---
output:
  word_document:
    reference_docx: myref.docx
bibliography: MyBib.bib
csl: Evolution.csl

@@

a. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
b. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
c. Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands
d. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
e. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles

@@

###Abstract
Understanding the links between early-life experiences and later life survival is fundamental to our understanding of life-history evolution. Telomeres - protective caps on the ends of chromosomes - shorten in response to early-life stress, and early-life telomere shortening is correlated with reduced later-life survival. As telomeres are a biomarker of the impact of early-life costs, they may provide a link between early-life environmental variation and later-life fitness. We studied how telomere dynamics link spatiotemporal variation in early-life conditions to survival using a long-term study of the Seychelles warbler (*Acrocephalus sechellensis*). We show that early-life telomere dynamics vary markedly across cohorts, and that individuals born in breeding seasons with higher food availability have longer telomeres and reduced telomere shortening. Within cohorts, we found that telomere length in nestlings was negatively related to tarsus length (which likely represents nestling growth), and positively related to the number of helpers present in the natal territory. Finally, we found that longer telomeres and reduced telomere shortening rates in early life were associated with increased survival later in life. Our results show that telomeres may provide a link between early-life conditions and late-life survival in wild populations.

**Keywords:** Life-history; Seychelles warbler; Senescence; Survival; Telomeres

@@

###Introduction
Being subject to favourable environmental conditions during development and growth can confer fitness advantages later in life (so called 'silver spoon effects') [@Grafen1988; @Monaghan2008]. Adult fitness can be affected by a range of early-life experiences, including variation in habitat quality and resource availability [@Madsen2000; @VandePol2006; @Hayward2013], population density [@Nussey2007;@Douhard2013] and natural or anthropogenic environmental disturbance [@Reid2003; @Cartwright2014]. Understanding the causes and consequences of these early-life experiences is key to understanding many ecological and evolutionary  processes, including  patterns of natural and sexual selection,  population growth rates and even local extinction [@Coulson2001; @Roach2014]. Understanding silver spoon effects is therefore of central interest to ecologists, evolutionary biologists and conservationists.

While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness. Because the phenotypic consequences of an adverse environment can depend on an individual's initial  condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006], the later-life consequences of a good or poor start are expected to vary among individuals, populations and species. Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships between the early-life conditions that are measured and adult phenotypes. Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009]. To better understand the later-life consequences of early-life experiences, we therefore need to understand, or at least be able to measure, how the environment differentially affects individuals within a population.

A key process that has been associated with adult survival is the rate at which telomere repeats are lost. Telomeres are protective caps on the ends of chromosomes which shorten with age [@Monaghan2006], and in response to oxidative stress levels which can be elevated  due to environmental factors [@VonZglinicki2002]. When telomeres reach a critical shortness cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003]. These links between telomere length and senescence have inspired a great deal of recent research into telomere ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014]. While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Barrett2011; @Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, retaining signatures of ecological stress that are otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].

There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life [e.g. @Frenck1998; @Haussmann2003], and that the extent of this telomere shortening is influenced by the conditions experienced early in life [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015]. Importantly, early-life telomere dynamics are related to both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and to other parameters such as cognition [@Nettle2015a]. However, the link between early-life conditions and late-life survival is limited to a few species, and thus poorly understood. Moreover, how early-life telomere dynamics vary over spatial and temporal scales is not known.

The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent model system for studying senescence in the wild [reviewed in @Hammers2015]. Due to the isolated nature of the study population and intensive field monitoring, we have excellent survival data and sampling material spanning many years (see methods, below). Ecological conditions and population density on Cousin are variable over space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a]. Seychelles warblers remain on their natal territories for at least six months, and variation in natal territory quality has been linked to changes in levels of oxidative stress [@VandeCrommenacker2011a]. However, neither natal territory quality nor natal local density appear to be linked to early-life or adult survival [@Brouwer2006; @Hammers2013]. Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordinates) in the natal territory has been associated with increased survival later in life [@Brouwer2012]. Lastly, we have an established protocol for assessing absolute telomere length in this species [@Barrett2012], and telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost [@Barrett2013a]. Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced early in life, and to assess the later life consequences of early-life conditions.

@@

Telomere lengths were measured for `r nrow(FlSA)+nrow(chicks)+nrow(Loss)` samples. Of these, `r nrow(FlSA)+nrow(chicks)` were taken cross-sectionally from birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`; `r nrow(chicks)` were nestlings less than one month old (chicks), while the remaining `r nrow(FlSA)` were fledglings and subadults aged between two and twelve months (hereafter referred to as fledglings). For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with an additional sample taken within two years of the original catch.  We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay  [full details in @Barrett2012].

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. Our sampling regime covers `r max(juv$LayYear)-min(juv$LayYear)` years, and temporal variation in rainfall and food availability on Cousin over this period is expected to result in variation  in a range of social and environmental variables [@Brouwer2012]. This makes it difficult to disentangle how spatial and temporal processes differentially affect early-life telomere dynamics. To overcome this, we analysed how variation in early-life conditions was related to variation in early-life telomere length separately among and within breeding seasons.

First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated average telomere length among all birds born in a given breeding season (i.e. a cohort), excluding cohorts where we had samples from $\leq$ 5 birds (n = `r nrow(juvseason)` cohorts). We tested for a difference in telomere length among cohorts using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among cohorts is explained by temporal variation in insect abundance.

@@

For individuals with longitudinal data (n = `r nrow(Loss)`) we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events. To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013]. Using this dataset we repeated the above analyses of among and within cohort telomere dynamics, replacing telomere length with telomere loss as the response variable.

Finally, we analysed the effects of telomere dynamics on survival, again by looking at variation both within and among cohorts. First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean telomere length and rate of loss of each cohort with the mean lifespan of all individuals in that cohort, using linear regression. We then used parametric survival analysis to test whether individual-level telomere length and rate of loss were related to to survival. Again, we mean-centred variables within cohorts to eliminate between-season temporal variation in telomeres and survival. The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values. As explanatory variables of survival we included early-life telomere length/loss (we ran a separate analysis for each, using the cross-sectional and longitudinal data, respectively), plus the factors identified as being related to telomere length in the previous analysis. With this approach we aimed to assess the effect of early-life telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere dynamics.

@@

#####Temporal variation in early-life telomere dynamics
Early-life telomere length varied significantly among breeding seasons (one-way ANOVA, F = `r round2(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~factor(FieldPeriodID),data = juv))[1,5])`), and this variation in average telomere length over breeding seasons was positively related to island-wide food availability (linear regression, R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data = juvseason,na.action=na.exclude))[1,5])`; Fig. 2A). Running the regression of cohort-level telomere length and food availability separately for each age class revealed a positive, significant relationship in fledglings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=flseason,na.action=na.exclude))[1,5])`), but no relationship in nestlings (R^2^ = `r round2(summary(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,4])`; *P* = `r round3(anova(lm(Insect~TLKBmean,data=chickseason,na.action=na.exclude))[1,5])`), indicating selective disappearance.

@@

#####Spatial variation in early-life telomere dynamics
The top model explaining within-season variation in early-life telomere length contained age class, tarsus length and the number of helpers present in the natal territory, as well as interactions between age class and both tarsus length and the number of helpers (Table S1). This model was much better supported than the null model ($\Delta$AICc = `r round2(AICc(lm(cenTL~1,data=juvall))-AICc(lm(cenTL~Fledged*(cenTarsus+cenHelper),data=juvall)))`). Both tarsus length and the number of helpers had high relative importance in the top model set (Fig. 3A), and the model containing these variables explained a reasonable amount of variation in telomere length (R^2^ = `r round2(summary(lm(cenTL~Ageclass*(cenTarsus+cenHelper),data=juvall))$r.squared)`). Telomere length was positively related to the number of helpers in nestlings, but not in fledglings (Fig. 3B). Tarsus length was negatively related to telomere length in nestlings, but not in fledglings (Fig. 3C).

For the subset of individuals with longitudinal data, we tested whether the factors previously identified as being related to early-life telomere length (i.e. tarsus length and number of helpers) were also associated with differences in rates of early-life telomere loss. Due to a limited sample size, we considered all age classes together for this analysis. We found that none of these variables were associated with differences in telomere loss (all *P* > 0.05).

#####early-life telomere dynamics and survival
At the population-level, there was no relationship between telomere length and lifespan (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig. 4A). Testing this relationship separately for each age class revealed that there was no age-specific relationship between cohort-level telomere length and lifespan (both *P* > 0.2). There was a negative relationship between population-level telomere loss and lifespan, although this was marginally non-significant (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig. 4B).

For the individual-based survival analysis, a log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution (Weibull) = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`). A survival model including all first year birds showed that early-life telomere length had a positive effect on survival, but this relationship was marginally non-significant (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`). Running this analysis among age classes revealed differential survival effects. Nestling telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; Fig. 5A), but telomere length at the fledgling stage was significantly related to survival (estimate = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`). This effect was positive, with longer telomeres at the fledgling stage associated with increased survival later in life (Fig. 6B). Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3], lessthan = F)`).

@@

Our finding that food availability is related to early-life telomere length and rate of shortening adds to a growing body of evidence from humans and wild animals showing that the natal environment can have pronounced impacts on early-life telomere dynamics [reviewed in @Price2013; @Monaghan2014].  However, very few studies have shown that temporal variation in environmental conditions affects telomere dynamics in natural populations, and the studies that have done so are limited to just two seasons [@Mizutani2013; @Watson2015]. The long-term Seychelles warbler dataset has allowed us, using many more cohorts than has been used in other studies, to show that temporal variation in environmental conditions does indeed affect telomere dynamics, and that this effect can be seen at the population level. Interestingly, in the Seychelles warbler annual variation in food availability is not directly linked to survival [@Brouwer2006; @Hammers2013]. It has been suggested that telomeres act as biomarkers of cost in natural populations, as they are able to capture individual and cohort-level variability in the costs experienced as a result of environmental pressures [@Asghar2015]. Our results confirm that telomeres can detect hidden costs in natural populations that are not detectable using life-history and environmental data alone

### 2015-10-05 Manuscript/SW_early_life_telomeres_SOM.Rmd
Kat B edits
HLD edits. Abstract and citation style changed to Evolution

###Supplementary information for Spurgin *et al.* "Telomeres reveal complex silver spoon effects in a wild population"

@@

**Table S1** Model selection table for analysis of factors affecting early life telomere length in Seychelles warbler chicks. Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here.

>>>>

###Supplementary information for Spurgin *et al.* "Telomeres reveal silver spoon effects in a wild population"

@@

**Table S1** Model selection table for analysis of factors affecting early life telomere length in Seychelles warblers. Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here. TQ = territory quality.
